\pkg{cv}: An \proglang{R} Package for Cross-Validating Regression Models

### 2025-05-09 scratch/JSS-submission/Final_submission/jss5535.Rmd
edit .Rmd, .tex, and update .pdf files

[^log-trans]: Although it serves to illustrate the use of CV, a polynomial is not the best choice here. Consider, for example, the scatterplot for log-transformed \code{mpg} and \code{horsepower}, produced by \code{plot(mpg $\sim$ horsepower, data=Auto, log="xy")} (execution of which is left to the reader). We revisit the \code{Auto} data in Section \ref{cross-validating-model-specification}.

Figure \ref{fig:mpg-horsepower-polynomials} (b) shows two measures of estimated (squared) error as a function of polynomial-regression degree: The mean-squared error ("MSE"), defined as $\mathsf{MSE} = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2$, and the usual residual variance, defined as $\widehat{\sigma}^2 = \frac{1}{n - p - 1} \sum_{i=1}^n (y_i - \widehat{y}_i)^2$. The former necessarily declines with $p$ (or, more strictly, can't increase with $p$), while the latter gets slightly larger for the largest values of $p$, with the "best" value, by a small margin, for $p = 7$.

@@

Using data like these, a researcher may attempt to obtain an estimate of the within-patient effect of dosage by fitting mixed models with a random intercept. We will illustrate how this approach results in estimates that are highly sensitive to the relative variation at the two levels of the mixed model by considering four models, all of which include a random intercept.  The first three models are sequentially nested in their fixed effects: (1) \code{~ 1}, intercept only; (2) \code{~ 1 + x}, intercept and effect of \code{x}; and (3) \code{~ 1 + x + xm}, intercept, effect of \code{x}, and a contextual variable, \code{xm}, consisting of the within-patient mean of \code{x}. The final model, (4) \code{~ 1 + I(x - xm)}, uses an intercept and the centered-within-group variable \code{x - xm}. We thus fit four models to four data sets for a total of 16 models.

@@

In Section \ref{comparing-competing-models}, following @JamesEtAl:2021[Secs. 5.1, 5.3], we fit polynomial regressions up to degree 10 to the relationship of \code{mpg} to \code{horsepower} for the \code{Auto} data, saving the 10 models, named \code{m.1} through \code{m.10}, in the list \code{mlist}. We then used \code{cv()} to compare the cross-validated MSE for the 10 models, discovering that the 7th degree polynomial had the smallest MSE (by a small margin).

@@

* \code{age}, in years.
* \code{econ.cond.national} and \code{econ.cond.household}, the respondent's ratings of the state of the economy, on 1 to 5 scales.
* \code{Blair}, \code{Hague}, and \code{Kennedy}, ratings of the leaders of the Labour, Conservative, and Liberal Democratic parties, on 1 to 5 scales.
* \code{Europe}, an 11-point scale on attitude towards European integration, with high scores representing "Euro-skepticism."
* \code{political.knowledge}, knowledge of the parties' positions on European integration, with scores from 0 to 3.
* \code{gender}, \code{"female"} or \code{"male"}.

@@

Figure \ref{fig:BEPS-plot} shows an "effect plot," using the the \pkg{effects} package [@FoxWeisberg:2019] to visualize the \code{Europe} $\times$ \code{political.knowledge} interaction in a "stacked-area" graph:
```{r BEPS-plot, fig.width=9, fig.height=5}
#| fig.cap = "Effect plot for the interaction between attitude towards European integration and political knowledge in the multinomial logit model fit to voting data from the 2001 British Election Panel Study, using party colors."
plot(effects::Effect(c("Europe", "political.knowledge"), m.beps,
            xlevels = list(Europe = 1:11, political.knowledge = 0:3),
            fixed.predictors = list(given.values = c(gendermale = 0.5))),
     lines = list(col = c("blue", "red", "orange")),
     axes = list(x = list(rug = FALSE), y = list(style = "stacked")))
```
As political knowledge increases, voters tend to align their votes more closely with the party positions on European integration: The Conservative Party was relatively Euro-skeptic, while Labour and the Liberal Democrats were more supportive of the UK's participation in the EU.

@@

The approach taken by \pkg{scikit-learn} for independent observations is similar to that in the default \code{cv()} method in the \pkg{cv} package. The approach for hierarchical data is similar, but not identical, to that in \pkg{cv} (as described in Section \ref{cross-validating-mixed-effects-models}). Our \pkg{cv} package doesn't currently provide for cross-validating time-series models, but we anticipate that the next version of the package will.

>>>>

[^log-trans]: Although it serves to illustrate the use of CV, a polynomial is not the best choice here. Consider, for example, the scatterplot for log-transformed \code{mpg} and \code{horsepower}, produced by \code{plot(mpg $\sim$ horsepower, data = Auto, log = "xy")} (execution of which is left to the reader). We revisit the \code{Auto} data in Section \ref{cross-validating-model-specification}.

Figure \ref{fig:mpg-horsepower-polynomials} (b) shows two measures of estimated (squared) error as a function of polynomial-regression degree: The mean-squared error ("MSE"), defined as $\mathsf{MSE} = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2$, and the usual residual variance, defined as $\widehat{\sigma}^2 = \frac{1}{n - p - 1} \sum_{i=1}^n (y_i - \widehat{y}_i)^2$. Here $y_i$ is the response and $\widehat{y}_i$ the fitted value for the $i$th case. The former necessarily declines with $p$ (or, more strictly, can't increase with $p$), while the latter gets slightly larger for the largest values of $p$, with the "best" value, by a small margin, for $p = 7$.

@@

Using data like these, a researcher may attempt to obtain an estimate of the within-patient effect of dosage by fitting mixed models with a random intercept. We will illustrate how this approach results in estimates that are highly sensitive to the relative variation at the two levels of the mixed model by considering four models, all of which include a random intercept.  The first three models are sequentially nested in their fixed effects: (1) \code{~ 1}, intercept only; (2) \code{~ 1 + x}, intercept and effect of \code{x}; and (3) \code{~ 1 + x + xm}, intercept, effect of \code{x}, and a contextual variable, \code{xm}, consisting of the within-patient mean of \code{x}. The final model, (4) \code{~ 1 + I(x - xm)}, uses an intercept and the centered-within-group variable \code{x - xm}. We thus fit four models to each of the four data sets for a total of 16 models.

@@

In Section \ref{comparing-competing-models}, following @JamesEtAl:2021[Secs. 5.1, 5.3], we fit polynomial regressions up to degree 10 to the relationship of \code{mpg} to \code{horsepower} for the \code{Auto} data, saving the 10 models, named \code{m.1} through \code{m.10}, in the object \code{mlist}. We then used \code{cv()} to compare the cross-validated MSE for the 10 models, discovering that the 7th degree polynomial had the smallest MSE (by a small margin).

@@

* \code{age}, in years.
* \code{econ.cond.national} and \code{econ.cond.household}, the respondent's ratings of the state of the economy, on 1 to 5 scales.
* \code{Blair}, \code{Hague}, and \code{Kennedy}, ratings of the leaders of the Labour, Conservative, and Liberal Democratic parties, on 1 to 5 scales.
* \code{Europe}, an 11-point scale for attitude towards European integration, with high scores representing "Euro-skepticism."
* \code{political.knowledge}, knowledge of the parties' positions on European integration, with scores from 0 to 3.
* \code{gender}, \code{"female"} or \code{"male"}.

@@

Figure \ref{fig:BEPS-plot} shows an "effect plot," using the the \pkg{effects} package [@FoxWeisberg:2019] to visualize the \code{Europe} $\times$ \code{political.knowledge} interaction in a "stacked-area" graph:
```{r BEPS-plot, fig.width=9, fig.height=5}
#| fig.cap = "Effect plot for the interaction between attitude towards European integration and political knowledge in the multinomial logit model fit to voting data from the 2001 British Election Panel Study, using party colors."
plot(effects::Effect(c("Europe", "political.knowledge"), m.beps,
            xlevels = list(Europe = 1:11, political.knowledge = 0:3),
            fixed.predictors = list(given.values = c(gendermale = 0.5))),
     lines = list(col = c("blue", "red", "orange")),
     axes = list(x = list(rug = FALSE), y = list(style = "stacked")))
```
As political knowledge increased, voters tended to align their votes more closely with the party positions on European integration: The Conservative Party was relatively Euro-skeptic, while Labour and the Liberal Democrats were more supportive of the UK's participation in the EU.

@@

The approach taken by \pkg{scikit-learn} for independent observations is similar to that in the default \code{cv()} method in the \pkg{cv} package. The approach for hierarchical data is similar, but not identical, to that in \pkg{cv} (as described in Section \ref{cross-validating-mixed-effects-models}). Our \pkg{cv} package doesn't currently provide for cross-validating time-series models, but we anticipate that a future version of the package will.
